Further Evidence of Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2022-03-21 src/ms.tex
adding Gaia

As discussed in \S\ref{subsec:models} and \S\ref{subsec:rossby}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}. Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is not biased against quiet, unspotted stars and is more likely to contain pile-up stars that our sample may have missed. It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge.

@@

We note that the \rocrit values we found in \S\ref{subsec:rossby} are smaller than the values of $\sim$2 found by \citet{vanSaders2016,vanSaders2019}, however, these results are sensitive to different \teff scales and different prescriptions for \taucz. We explored the \taucz formulae of \citet{BarnesKim2010} and \citet{Landin2010}, but these produce larger \taucz at fixed \teff than the \citet{Gunn1998} relation, which exacerbates the \rocrit mismatch mentioned above.

@@

\begin{acknowledgments}
We thank the anonymous referee for a thoughtful review, Rodrigo Luger for helpful discussions and assistance with the \texttt{showyourwork!} package, Adrian Price-Whelan for providing the APOGEE--Kepler cross-match catalog, and Karl Jaehnig for plotting advice. It is a pleasure to thank the Stars Group at the American Museum of Natural History, and the Astronomical Data Group at the Flatiron Institute for helpful discussions.

>>>>

As discussed in \S\ref{subsec:rossby} and \S\ref{subsec:models}, the existence of a pile-up of stars with a constant Rossby number is consistent with expectations from the WMB model of \citet{vanSaders2016, vanSaders2019}. The location of the long-period pile-up coincides closely with the cluster of main-sequence stars with asteroseismic rotation rates determined by \citet{Hall2021} and we suggest the two features are one and the same. Offsets in both observational and theoretical \teff scales between different studies can explain the modest discrepancies between the asteroseismic sample and the samples considered here, as well as discrepancies in the \rocrit values inferred here and those proposed in \citet{vanSaders2019}.

Furthermore, we note that there are important differences between the samples considered here and the \citet{Hall2021} asteroseismic sample. The present samples have rotation periods detected from photometric variations and, for a survey of finite baseline and sensitivity like Kepler, photometric rotation periods are harder to detect for slower rotators, smaller amplitude variations, and more stochastic variability patterns. Exacerbating the problem of detection, slower rotation tends to be accompanied by smaller-scale and less periodic variability. The asteroseismic sample, by comparison, is not biased against quiet, unspotted stars and is more likely to contain pile-up stars that our sample may have missed.

It is evident from Figure~\ref{fig:gap} that the long-period pile-up also coincides with a steep gradient in variability amplitude, such that a hidden population of pile-up stars may lie just beyond the edge of detectability. This effect, in addition to the \teff offsets mentioned above, can further bias our inference of \rocrit to lower values. Regardless of the biases mentioned above and the true value of \rocrit, the sample considered clearly indicates that the long-period pile-up is apparent for stars with \teff~$\gtrsim$~\teffmin with Rossby numbers lower than those associated with a detection edge.

@@

We note that the \rocrit values we found in \S\ref{subsec:rossby} are smaller than the values of $\sim$2 found by \citet{vanSaders2016,vanSaders2019}, however, these results are sensitive to different \teff scales and different prescriptions for \taucz. We explored the \taucz formulae of \citet{BarnesKim2010} and \citet{Landin2010}, but these produce larger \taucz at fixed \teff than the \citet{Gunn1998} relation, which exacerbates the \rocrit mismatch mentioned above. \edit1{As shown in \citet{See2021}, the \taucz relation of \citet{Amard2019} also lies above the \citet{Gunn1998} relation, so adopting the newer \citet{Amard2019} relation would also not alleviate the \rocrit discrepancy.}

@@

\section{Gaia}\label{app:gaia}
\edit1{Given that the long-period pile-up was previously obscured by imprecise \teff measurements, we explored whether the feature could be recovered using the high precision photometric colors provided by the Gaia mission \citep{Gaia2016}. We cross-matched the \citet{McQuillan2014} sample with Gaia EDR3 \citep{GaiaEDR3}. We then compared the ($G_{BP}-G_{RP}$)–\prot distribution with constant Rossby curves. For this exercise, we used the empirically calibrated \taucz relation of \citet{Corsaro2021}, who presented \taucz as a quadratic function of Gaia $G_{BP}-G_{RP}$ color using the Kepler LEGACY asteroseismic sample as calibrators.}

\begin{acknowledgments}
We thank the anonymous referee for a thoughtful review, Rodrigo Luger for helpful discussions and assistance with the \texttt{showyourwork!} package, Adrian Price-Whelan for providing the APOGEE--Kepler cross-match catalog, Karl Jaehnig for plotting advice, and Matteo Cantiello for helpful discussions. It is a pleasure to thank the Stars Group at the American Museum of Natural History, and the Astronomical Data Group at the Flatiron Institute for helpful discussions.
